---
title: "coextinction_cascades"
author: "Alejandro Santillana"
date: "2024-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(bipartite)
library(fossil)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(networkD3)
library(terra)
library(lme4)
library(ggspatial)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(viridis)
```

second.extinct calculates the consequences of removing a species from a bipartite network. With plotting and slope-estimation functionality.

Details: The procedure of this function is simple. For example imagine the web to represent a pollination web, in which pollinators die one by one. Set all entries of a column to zero, see how may rows are now also all-zero (i.e. species that are now not pollinated any more), and count these as secondary extinctions of the primary exterminated pollinator.

Internally, each extermination is achieved by a call to extinction, followed by a call to empty, which counts the number of all-zero columns and rows.

Although written for pollination webs (hence the nomenclature of pollinator and plant), it can be similarly applied to other types of bipartite networks. It is called by networklevel.
Usage: second.extinct(web, participant = "higher", method = "abun", nrep = 10, details = FALSE, ext.row=NULL, ext.col=NULL)

Arguments:
  method: Random deletion of a species (random); according to its abundance, with least abundant going extinct first (abundance; default) or "degree" to build a sequence from the best-to-least connected species. This is the most extreme case, where the most generalist species goes extinct first (see Memmott et al. 1998). (Note that method="abundance" does not work with participant="both"; in this case a random sequence of species from both trophic levels is generated.) external will use the externally provided vector to determine extinction sequence.
  ext.row: Optional vector giving the sequence in which lower-level species are to be deleted.
  ext.col: Optional vector giving the sequence in which higher-level species are to be deleted.
  
Note: networklevel calls second.extinct; extinction and empty are internal helper functions, and slope.bipartite calculates extinction slopes from the output of second.extinct.

```{r second.extinct}
# ?second.extinct
# ?slope.bipartite
# ?robustness
```

bipartite function "networklevel" gives outputs:
 $‘extinction slope lower trophic level‘ [1] 2.286152 
 $‘extinction slope higher trophic level‘ [1] 2.9211

Extinction slopes are hyperbolic fits to a simulated extinction sequence of the network, which causes secondary extinctions in the other trophic level (from Memmott 2004).

slope.bipartite fits a hyperbolic function to the extinction simulation of "second.extinct."
This function scales extinction sequences to values between 0 and 1 for each participant. 
The x-axis of the graph features the proportion of exterminated participants, while the y-axis depicts the proportion of secondary extinctions. Since these curves usually follow a hyperbolic function (see examples in Memmott et al. 2004), this is fitted to the data.

## Sky Islands Networks

``` {r load sky islands networks}
load("../skyIslands/data/networks/Yr_PlantPollinator_Bees.RData") #yearly networks for bees
# load("../skyIslands/data/networks/YrSR_PlantPollinator_Bees.RData") #survey round networks for bees

```

```{r robustness}
# networklevel(nets[["CH.2018"]])
# 
# trial <- second.extinct(nets[["CH.2018"]], participant = "higher", method = "abun", nrep = 500, details = FALSE)
# trial_slope <- slope.bipartite(trial, plot.it = TRUE)

# Create an empty data frame to store the results
robustness_results <- data.frame(Network = character(), Site = character(), Year = integer(), Robustness = numeric(), stringsAsFactors = FALSE)

# Loop through each network in the list `nets`
for (net_name in names(nets)) {
  # Extract the network
  network <- nets[[net_name]]
  
  # Run extinction simulation
  network_ext <- second.extinct(network, participant = "higher", method = "abun", nrep = 50, details = FALSE)
  
  # Calculate the robustness (area under second.extinct curve)
  network_robustness <- robustness(network_ext)
  
  # Extract Site and Year from the network name
  site_year <- unlist(strsplit(net_name, "\\.")) # Split by "."
  site <- site_year[1]
  year <- as.integer(site_year[2])
  
  # Append the results to the data frame
  robustness_results <- rbind(robustness_results, data.frame(Network = net_name, Site = site, Year = year, Robustness = network_robustness))
}

# Print the resulting data frame
print(robustness_results)

```

Import Precipitation Rasters from PRISM

``` {r import rasters}
precip_2011_06_raster <- rast("PRISM_summer_precip/2011/PRISM_ppt_stable_4kmM3_201106_bil.bil")
precip_2011_07_raster <- rast("PRISM_summer_precip/2011/PRISM_ppt_stable_4kmM3_201107_bil.bil")
precip_2011_08_raster <- rast("PRISM_summer_precip/2011/PRISM_ppt_stable_4kmM3_201108_bil.bil")
precip_2011_09_raster <- rast("PRISM_summer_precip/2011/PRISM_ppt_stable_4kmM3_201109_bil.bil")

precip_2016_06_raster <- rast("PRISM_summer_precip/2016/PRISM_ppt_stable_4kmM3_201606_bil.bil")
precip_2016_07_raster <- rast("PRISM_summer_precip/2016/PRISM_ppt_stable_4kmM3_201607_bil.bil")
precip_2016_08_raster <- rast("PRISM_summer_precip/2016/PRISM_ppt_stable_4kmM3_201608_bil.bil")
precip_2016_09_raster <- rast("PRISM_summer_precip/2016/PRISM_ppt_stable_4kmM3_201609_bil.bil")

precip_2017_06_raster <- rast("PRISM_summer_precip/2017/PRISM_ppt_stable_4kmM3_201706_bil.bil")
precip_2017_07_raster <- rast("PRISM_summer_precip/2017/PRISM_ppt_stable_4kmM3_201707_bil.bil")
precip_2017_08_raster <- rast("PRISM_summer_precip/2017/PRISM_ppt_stable_4kmM3_201708_bil.bil")
precip_2017_09_raster <- rast("PRISM_summer_precip/2017/PRISM_ppt_stable_4kmM3_201709_bil.bil")

precip_2020_06_raster <- rast("PRISM_summer_precip/2020/PRISM_ppt_stable_4kmM3_202006_bil.bil")
precip_2020_07_raster <- rast("PRISM_summer_precip/2020/PRISM_ppt_stable_4kmM3_202007_bil.bil")
precip_2020_08_raster <- rast("PRISM_summer_precip/2020/PRISM_ppt_stable_4kmM3_202008_bil.bil")
precip_2020_09_raster <- rast("PRISM_summer_precip/2020/PRISM_ppt_stable_4kmM3_202009_bil.bil")

precip_2021_06_raster <- rast("PRISM_summer_precip/2021/PRISM_ppt_stable_4kmM3_202106_bil.bil")
precip_2021_07_raster <- rast("PRISM_summer_precip/2021/PRISM_ppt_stable_4kmM3_202107_bil.bil")
precip_2021_08_raster <- rast("PRISM_summer_precip/2021/PRISM_ppt_stable_4kmM3_202108_bil.bil")
precip_2021_09_raster <- rast("PRISM_summer_precip/2021/PRISM_ppt_stable_4kmM3_202109_bil.bil")

# Define the years and file paths
years <- c("2011", "2016", "2017", "2020", "2021")
months <- c("06", "07", "08", "09")
base_path <- "PRISM_summer_precip"

# Initialize a list to store the summed rasters for each year
summer_rasters <- list()

# Loop through each year
for (year in years) {
  
  # Initialize an empty list to hold monthly rasters
  monthly_rasters <- list()
  
  # Loop through each month
  for (month in months) {
    # Construct the file path
    file_path <- file.path(base_path, year, paste0("PRISM_ppt_stable_4kmM3_", year, month, "_bil.bil"))
    
    # Load the raster and append to the list
    monthly_rasters[[month]] <- rast(file_path)
  }
  
  # Stack and sum the monthly rasters
  stacked_rasters <- rast(monthly_rasters)
  summer_raster <- sum(stacked_rasters, na.rm = TRUE)
  
  # Save the raster to the list
  summer_rasters[[year]] <- summer_raster
  
  # Optionally, save the raster to a file
  #writeRaster(summer_raster, paste0("precip_", year, "_raster.tif"), overwrite = TRUE)
  
  # Print a message for progress tracking
  print(paste("Processed year:", year))
}

precip_2011_raster <- summer_rasters[["2011"]]
precip_2016_raster <- summer_rasters[["2016"]]
precip_2017_raster <- summer_rasters[["2017"]]
precip_2020_raster <- summer_rasters[["2020"]]
precip_2021_raster <- summer_rasters[["2021"]]

# Identify variables with the pattern "precip_****_**_raster"
monthly_rasters <- ls(pattern = "^precip_\\d{4}_\\d{2}_raster$")
rm(list = monthly_rasters)

```

Import Sites Shapefile and Change CRS

``` {r import site shapefile}
sites_shp <- vect("Sites/sites.shp")
print(crs(sites_shp))
print(crs(precip_2011_raster))
sites_shp <- project(sites_shp, crs(precip_2011_raster))
```

Extract Precipitation Data for Sites
``` {r for loop}
years <- c(2011, 2016, 2017, 2020, 2021)
results_list <- list()

for (year in years) {
  # Load the raster for the year
  raster <- rast(paste0("PRISM_precip/PRISM_ppt_stable_4kmM3_", year, "_bil.bil")) 
  
  # Extract precipitation values for subsites
  precip_values <- extract(raster, sites_shp)
  
  # Combine extracted values with site shapefile data
  subsite_data <- as.data.frame(sites_shp) # Convert shapefile to a data frame
  subsite_data$Precipitation <- precip_values[, 2] # Add extracted values
  subsite_data$Year <- year # Add the year
  
  # Group by Site and Year, then calculate mean precipitation
  site_data <- subsite_data %>%
    group_by(Site, Year) %>%
    summarize(Mean_Precipitation = mean(Precipitation, na.rm = TRUE), .groups = "drop")
  
  # Store the yearly averaged data in the results list
  results_list[[as.character(year)]] <- site_data
}

# Combine all yearly site-level results into a single data frame
all_precip_data <- do.call(rbind, results_list)
```

Merge with Robustness Results
``` {r merge precip with robustness}
# Add one year to the precipitation year to align with robustness data
all_precip_data <- all_precip_data %>%
  mutate(Year = Year + 1) 

final_data <- merge(robustness_results, all_precip_data, by = c("Site", "Year"))
final_data <- final_data %>%
  filter(!Site %in% c("SS", "UK", "VC"))
```

Fit a Linear Mixed-effects Model
``` {r model}
model <- lmer(Robustness ~ Mean_Precipitation + Year + (1 | Site), data = final_data)

summary(model)
```

Visualize
``` {r visualization}
# # Plot Robustness vs. Mean_Precipitation by year
# ggplot(final_data, aes(x = Mean_Precipitation, y = Robustness, color = as.factor(Year))) +
#   geom_point(size = 3) +
#   geom_smooth(method = "lm", se = FALSE) +
#   labs(title = "Network Robustness vs. Summer Precipitation by Year",
#        x = "Mean Precipitation (previous summer) (mm)",
#        y = "Network Robustness",
#        color = "Year") +
#   theme_minimal()
  
ggplot(final_data, aes(x = Mean_Precipitation, y = Robustness, color = Site, shape = as.factor(Year))) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, aes(group = 1), color = "black", linetype = "dashed") +
  labs(
    title = "Network Robustness by Precipitation and Site",
    subtitle = "Mixed-effects model: small site-level variance, negative trend with precipitation",
    x = "Mean Precipitation (previous year) (mm)",
    y = "Network Robustness",
    color = "Site",
    shape = "Year"
  ) +
  annotate("text", x = 500, y = 0.2, label = "Slope = -0.00028\nSite variance = 0.0052\np-value = 0.086", size = 4, color = "blue") +
  theme_minimal()



# # Plot Robustness vs Mean_Precipitation by site
# ggplot(final_data, aes(x = Mean_Precipitation, y = Robustness)) +
#   geom_point(aes(color = Site), size = 3) +       # Points colored by Site
#   geom_smooth(method = "lm", se = TRUE, color = "blue") + # Add linear regression line
#   theme_minimal() +
#   labs(title = "Robustness vs. Summer Precipitation",
#        x = "Mean Precipitation (previous summer) (mm)",
#        y = "Network Robustness")
```


Sankey Plot
``` {r sankey}
# makeSankeyPlot<- function(i){
#   print(i)
#   print(names(nets)[i])
#   data_long <- as.data.frame(nets[[i]]) %>%
#     rownames_to_column %>%
#     gather(key = 'key', value = 'value', -rowname) %>%
#     filter(value > 0)
#   colnames(data_long) <- c("source", "target", "value")
#   data_long$target <- paste(data_long$target, " ", sep="")
#   
#   ## From these flows we need to create a node data frame: it lists
#   ## every entities involved in the flow
#   nodes <- data.frame(name=c(as.character(data_long$source),
#                              as.character(data_long$target)) %>% unique())
#   
#   ## With networkD3, connection must be provided using id, not using
#   ## real name like in the links dataframe.. So we need to reformat
#   ## it.
#   data_long$IDsource=match(data_long$source, nodes$name)-1 
#   data_long$IDtarget=match(data_long$target, nodes$name)-1  
#   net.p <- sankeyNetwork(Links = data_long, Nodes = nodes,
#                          Source = "IDsource", Target = "IDtarget",
#                          Value = "value", NodeID = "name", 
#                          fontSize=18)
#   ## colourScale=rainbow(
#   ##   prod(dim(bee.nets[[i]]))))
#   saveNetwork(net.p, sprintf("figures/sankey_%s.html",
#                              gsub(" ", "", names(nets)[i])))
#   webshot(sprintf("figures/sankey_%s.html",
#                   gsub(" ", "", names(nets)[i])),
#           file=sprintf("figures/sankey_%s.jpeg",
#                        gsub(" ", "",
#                             names(nets)[i])),
#           vwidth = 650, vheight = 1400)
#   
# }
# lapply(1:length(nets), makeSankeyPlot)
```


Extinction Curve Plot
``` {r extinction curve}
trial <- second.extinct(nets[["CH.2018"]], participant = "higher", method = "abun", nrep = 1000, details = FALSE)
trial_slope <- slope.bipartite(trial, plot.it = TRUE, ylab="y")
```

Map of Sky Islands
``` {r sky islands map}
# Load the country or region boundary
world <- ne_countries(scale = "medium", returnclass = "sf")

# (Optional) Subset for specific countries if relevant
world <- world[world$name %in% c("United States of America"), ]

# Get U.S. state boundaries
states <- ne_states(country = "United States of America", returnclass = "sf")


sites <- st_as_sf(sites_shp)

ggplot() +
  # Add basemap
  geom_sf(data = world, fill = "lightgray", color = "white") +
  
  # Add your sites
  geom_sf(data = sites, color = "red", size = 3, shape = 21, fill = "yellow") +
  
  # Add state boundaries
  geom_sf(data = states, fill = NA, color = "black", linetype = "dashed") +
  
  # Add optional scalebar and north arrow
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl", which_north = "true", 
                         style = north_arrow_fancy_orienteering) +
  
  # Define map extent for the southwestern U.S.
  coord_sf(xlim = c(-115, -102), ylim = c(30, 38), expand = FALSE) +
  
  # Customize the appearance
  theme_minimal() +
  labs(title = "Sky Island Sites",
       subtitle = "Geographic distribution in the Southwestern U.S.",
       x = "Longitude", y = "Latitude")
```
Precipitation map
``` {r precip map}
# # Define a list of the rasters
# rasters <- list(
#   precip_2011 = precip_2011_raster,
#   precip_2016 = precip_2016_raster,
#   precip_2017 = precip_2017_raster,
#   precip_2020 = precip_2020_raster,
#   precip_2021 = precip_2021_raster
# )
# 
# # Define a list of corresponding years for labeling
# years <- c("2011", "2016", "2017", "2020", "2021")
# 
# # Calculate global minimum and maximum precipitation values across all rasters
# min_precip <- min(sapply(rasters, function(r) min(values(r), na.rm = TRUE)))
# max_precip <- max(sapply(rasters, function(r) max(values(r), na.rm = TRUE)))
# 
# # Loop through each raster and create a plot for each year
# for (i in 1:length(rasters)) {
#   
#   # Get the current raster and year
#   current_raster <- rasters[[i]]
#   current_year <- years[i]
#   
#   # Convert the current raster to a data.frame for ggplot
#   precip_df <- as.data.frame(current_raster, xy = TRUE, na.rm = TRUE)
#   
#   # Rename the third column to "precip"
#   colnames(precip_df)[3] <- "precip"
#   
#   # Apply log transformation to precipitation values (log(x + 1) to avoid log(0))
#   precip_df$precip_log <- log(precip_df$precip + 1)
#   
#   # Create the plot for the current year
#   plot <- ggplot() +
#     # Add the precipitation raster layer
#     geom_raster(data = precip_df, aes(x = x, y = y, fill = precip_log)) +
#     
#     # Add a color scale for the precipitation
#    scale_fill_viridis(option = "D", name = paste("Precipitation", year, "(log scale)"), direction = 1, 
#                       limits = c(log(min_precip + 1), log(max_precip + 1))) +
#     
#     # Add the site locations
#     geom_sf(data = sites, color = "red", size = 3, shape = 21, fill = "yellow") +
#     
#     # Customize map extent (adjust as needed)
#     coord_sf(xlim = c(-111, -105), ylim = c(31, 37), expand = FALSE) +
#     
#     # Customize the appearance
#     theme_minimal() +
#     labs(
#       title = paste("Sky Island Sites and Precipitation in", current_year, "(log-transformed)"),
#       subtitle = paste("Precipitation layer for the year", current_year),
#       x = "Longitude", 
#       y = "Latitude"
#     ) +
#     theme(legend.position = "right")
#   
#   # Print the plot for the current year
#   print(plot)
# }

# Define a list of the rasters
rasters <- list(
  precip_2011 = precip_2011_raster,
  precip_2016 = precip_2016_raster,
  precip_2017 = precip_2017_raster,
  precip_2020 = precip_2020_raster,
  precip_2021 = precip_2021_raster
)

# Define a list of corresponding years for labeling
years <- c("2011", "2016", "2017", "2020", "2021")

# Calculate global minimum and maximum precipitation values across all rasters
min_precip <- min(sapply(rasters, function(r) min(values(r), na.rm = TRUE)))
max_precip <- max(sapply(rasters, function(r) max(values(r), na.rm = TRUE)))

# Create an empty list to store all the data for ggplot
plot_data <- list()

# Loop through each raster and prepare data for ggplot
for (i in 1:length(rasters)) {
  
  # Get the current raster and year
  current_raster <- rasters[[i]]
  current_year <- years[i]
  
  # Convert the current raster to a data.frame for ggplot
  precip_df <- as.data.frame(current_raster, xy = TRUE, na.rm = TRUE)  # Remove NAs
  
  # Rename the third column to "precip"
  colnames(precip_df)[3] <- "precip"
  
  # Apply log transformation to precipitation values (log(x + 1) to avoid log(0))
  precip_df$precip_log <- log(precip_df$precip + 1)
  
  # Add the year to the data for faceting
  precip_df$year <- current_year
  
  # Store the data in the list
  plot_data[[i]] <- precip_df
}

# Combine all the data into a single data frame
plot_data_df <- do.call(rbind, plot_data)

# Create the plot with facet wrap
plot <- ggplot(plot_data_df) +
  # Add the precipitation raster layer
  geom_raster(aes(x = x, y = y, fill = precip_log)) +
  
  # Add a color scale for the precipitation with fixed limits for consistency
  scale_fill_viridis(option = "D", name = "Precipitation",
                     direction = 1) +
# limits = c(log(min_precip + 1), log(max_precip + 1))) +
  
  # Add the site locations
  geom_sf(data = sites, color = "black", size = 3, shape = 21, fill = "grey") +
  
  # Customize map extent (adjust as needed)
  coord_sf(xlim = c(-111, -105), ylim = c(31, 37), expand = FALSE) +
  
  # Customize the appearance
  theme_minimal() +
  labs(
    title = "Sky Island Sites and Precipitation",
    subtitle = "Summer precipitation layer for different years (log-transformed)",
    x = "Longitude", 
    y = "Latitude"
  ) +
  # Round axis labels
  scale_x_continuous(breaks = seq(-110, -106, by = 2), 
                     labels = function(x) round(x, 2)) +  # Adjust longitude rounding
  scale_y_continuous(breaks = seq(31, 37, by = 1), 
                     labels = function(y) round(y, 2)) +  # Adjust latitude rounding
  
  # Facet wrap by year
  facet_wrap(~ year, nrow=1)

# Print the plot
print(plot)


```
